---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r}
## Load dplyr package
library(dplyr)

## Load data to data table 
data <- tbl_df(read.csv("activity.csv", stringsAsFactors = FALSE))

## Convert Date column to date type
data$date <- as.Date(data$date, format = "%Y-%m-%d")
```

## What is mean total number of steps taken per day?

```{r}

## Group by date and then sum the total number of steps removing NA entries
totalSteps <- group_by(data, date) %>% 
    summarize_each(funs(sum(., na.rm = TRUE))) 

## Create a histogram of total number of steps
hist(totalSteps$steps, 
     col = "red", 
     xlab = "Total Steps", 
     main = "Total Number of Steps Taken Each Day")

## Calculate the mean and save to a variable
meanSteps <- mean(totalSteps$steps)

```

The mean number of steps taken each day is `r meanSteps`  

```{r}

## Calculate the median and save to a variable
medianSteps <- median(totalSteps$steps)

```

The median number of steps taken each day is `r medianSteps`  


## What is the average daily activity pattern?

```{r}

## Select the steps and interval columns, Group by interval and then get the average number of steps for each interval
avgStepsPerInterval <- select(data, steps, interval) %>% 
    group_by(interval) %>% 
    summarize_each(funs(mean(., na.rm = TRUE))) 

## Plot steps per interval
plot(avgStepsPerInterval$interval, 
     avgStepsPerInterval$steps, 
     type = "l", 
     xlab = "Interval", 
     ylab = "Average Number of Steps", 
     col="blue", 
     main = "Average Number of Steps by Interval") 

```

```{r}

maxActiveInterval <- avgStepsPerInterval[which.max(avgStepsPerInterval$steps),]$interval

```

The 5 minute interval that contains the maximum number of steps on average over all days is `r maxActiveInterval`   

## Imputing missing values

```{r}

## Get the number of rows with missing values
naRows <- nrow(data[!complete.cases(data),])

```

There are `r naRows` rows with missing values in the dataset.

```{r}

## Create a copy of the data frame
fullData <- data

## Fill missing steps values with the average for the interval
fullData$steps <- ifelse(is.na(fullData$steps), avgStepsPerInterval$steps[match(fullData$interval, avgStepsPerInterval$interval)], fullData$steps)

## Group by date and then sum the total number of steps
fullDataTotalSteps <- group_by(fullData, date) %>% 
    summarize_each(funs(sum(., na.rm = TRUE))) 

## Create a histogram of total number of steps
hist(fullDataTotalSteps$steps, 
     col = "red", 
     xlab = "Total Steps", 
     main = "Total Number of Steps Taken Each Day After Imputing Values")

## Calculate the mean and save to a variable
fullDataMeanSteps <- mean(fullDataTotalSteps$steps)

```

The mean number of steps taken each day after imputing values is `r fullDataMeanSteps`  

```{r}

## Calculate the median and save to a variable
fullDataMedianSteps <- median(fullDataTotalSteps$steps)

```

The median number of steps taken each day after imputing values is `r fullDataMedianSteps`  

These values differ from the estimates from the first part of the assignment.  

After imputing missing data the total daily number of steps skews to the average as expected.  


## Are there differences in activity patterns between weekdays and weekends?

```{r}

## Create the factor column weekPart and assign the appropriate value
fullData$weekPart <- factor(ifelse(weekdays(fullData$date) %in% c("Saturday", "Sunday"),"weekend", "weekday"))

## Select the steps interval, and weekPart columns, Group by interval and weekPart  then get the average number of steps for each interval per weekPart
avgStepsPerIntervalByWeekPart <- select(fullData, steps, interval, weekPart) %>% 
    group_by(interval, weekPart) %>% 
    summarize_each(funs(mean(., na.rm = TRUE))) 

library(lattice)

## Plot steps per interval and weekPart
xyplot(steps ~ interval | weekPart,
       data = avgStepsPerIntervalByWeekPart,
       type = "l", 
       xlab = "Interval", 
       ylab = "Average Number of Steps", 
       col = "blue", 
       layout=c(1,2),
       main = "Average Number of Steps by Interval and Week Part")

```